Cubical Types

### 2014-05-18 ring.tex
nice beautiful figures

\documentclass[authoryear,preprint]{sigplanconf}
\usepackage{amsmath}
\usepackage{listings} 
\usepackage{stmaryrd}
\usepackage{latexsym}
\usepackage{amssymb}
\usepackage{xcolor}
\usepackage{courier}
\usepackage{thmtools}
\usepackage{bbold}

\newtheorem{definition}{Definition}
\begin{document}
\special{papersize=8.5in,11in}
\setlength{\pdfpageheight}{\paperheight}
\setlength{\pdfpagewidth}{\paperwidth}

@@

%% double-blind reviewing...
\title{Negative Types}
\authorinfo{}{}{}
\maketitle

@@

Ignore function types for now. Finite algebraic types (constructed from 0, 1,
sum, and product) are positive types; they represent the program (as opposed
to the environment), the player (as opposed to the opponent), the producer
(as opposed to the consumer). Negative types represent the environment, the
opponent, the consumer, etc. There is a nice symmetry here but that is
difficult to see if the computational model has information
effects. Intuitively, if the computational infrastructure allows arbitrary
things to be discarded or duplicated then ``loans'' or ``debts'' must be
introduced with extreme care. However in settings where resources are
maintained, it is conceptually simpler to introduce ``debts'' and liberally
compute with ``positive'' and ``negative'' values knowing that every demand
for a value will be eventually matched up with a corresponding positive
flow.

Other than semantic elegance, the immediate significance is that we get
higher-order functions for free as objects of type $-a+b$ expressing the
demand for a value of type $a$ and the production of a value of type $b$.  We
can see this beautiful symmetry and the emergence of functions from symmetry
in linear logic. The geometry of interaction construction takes a fragment of
linear logic and builds higher-order functions. It however doesn't work for
full linear logic. The related Int construction does not work if we have two
functors. This turns out to be a long-standing open problem in algebraic
topology that was recently solved in the ``ring completion'' paper. So the
technical question is how to take computational infrastructure where
information is conserved and add these symmetries that lead to higher-order
functions without losing the product structure.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Conclusion}

>>>>

\documentclass[authoryear,preprint]{sigplanconf}
\usepackage{amsmath}
\usepackage{listings} 
\usepackage{stmaryrd}
\usepackage{latexsym}
\usepackage{amssymb}
\usepackage{xcolor}
\usepackage{courier}
\usepackage{thmtools}
\usepackage{bbold}
\usepackage{tikz}

\newcommand{\lolli}{\multimap} 
\newcommand{\cubt}{\mathbb{T}}
\newcommand{\nodet}[2]{\fcolorbox{black}{white}{$#1$}\fcolorbox{black}{gray!20}{$#2$}}

\begin{document}
\special{papersize=8.5in,11in}
\setlength{\pdfpageheight}{\paperheight}
\setlength{\pdfpagewidth}{\paperwidth}

@@

%% double-blind reviewing...
\title{Cubical Types}
\authorinfo{}{}{}
\maketitle

@@

Int construction of G construction is neat. As Neel explains, given
first-order types and feedback you get higher-order functions. But if you do
the construction on the additive structure, you lose the multiplicative
structure. It turns out that this is related to a deep open problem in
algebraic topology and homotopy theory that was recently solved. We
``translate'' that solution to a computational type-theoretic world. This has
evident connections to homotopy (type?) theory that remain to be
investigated.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{The Int Construction}

Explain in detail perhaps with Haskell embedding and type functions etc. The
key insight it to add ``negative'' types representing demand for values. This
is how you get functions.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{The Problem and the Intuitive Solution}

We lose the multiplicative structure. No evident way to define the
multiplication functor. Perhaps there is a clever way. But this turns out to
be a well-known open problem. Review the problem and tell the story from the
algebraic topology perspective.

Regular first-order types are viewed as 1-dimensional cubes. Generalize to
$n$-dimensional cubes.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{The Formal Construction}

We use $\tau$ for regular first-order types which include the empty type, the
unit type, and conventional sum and product types.  We use $\cubt$ for
general $n$-dimensional cubical types. An $n$-dimensional cubical type
$\nodet{\cubt_1}{\cubt_2}$ consists of a positive subspace $\cubt_1$ and a
negative subspace $\cubt_2$ (shaded in gray). Each of the subspaces are
cubical types of a lower dimension; the 1-dimensional case reduces to plain
first-order types. As we explain below, cubical types are closed under sums,
differences, products, and can express higher-order functions $\cubt_1 \lolli
\cubt_2$ as $(\ominus \cubt_1) \oplus \cubt_2$. The syntax of the types is
summarized below:
\[\begin{array}{rcl}
\tau &::=& 0 \alt 1 \alt \tau_1 + \tau_2 \alt \tau_1 * \tau_2 \\
\\
\cubt &::=& \tau \alt \nodet{\cubt_1}{\cubt_2} 
      \alt \cubt \oplus \cubt 
      \alt \cubt \otimes \cubt 
      \alt \ominus \cubt 
\end{array}\]
Examples of 1-dimensional cubical types include all the regular first-order
types $\tau$. A 2-dimensional cubical type looks like
$\nodet{\tau_1}{\tau_2}$ which intuitively corresponds to the difference
$\tau_1 - \tau_2$ of the two types. A 3-dimensional cubical type looks like
$\nodet{(\nodet{\tau_1}{\tau_2})}{(\nodet{\tau_3}{\tau_4})}$ which
intuitively corresponds to the iterated difference of types
$(\tau_1-\tau_2)-(\tau_3-\tau_4)$ where the successive colors from the root
encode the sign as follows: a color that is the same as its enclosing color
is positive and a color that differs from its enclosing color is negative. It
is possible for cubical types not be balanced in general but it is always
possible to extend the shorter subtree to an arbitrary depth using the fact
(to be justified later) that~$\tau$ is equivalent to $\nodet{\tau}{0}$.

We now explain how cubical types are closed under sum, difference, and
products. In the recursive definitions below, we assume that the types have
been extended to be balanced as needed:
\[\begin{array}{rcl}
\ominus~\tau &=& \tau \\
\ominus~\nodet{\cubt_1}{\cubt_2} &=& \nodet{\ominus~\cubt_2}{\ominus~\cubt_1} \\
\\
\tau_1 \oplus \tau_2 &=& \tau_1 + \tau_2 \\
(\nodet{\cubt_1}{\cubt_2}) \oplus (\nodet{\cubt_3}{\cubt_4}) &=& 
  \nodet{\cubt_1 \oplus \cubt_3}{\cubt_2 \oplus \cubt_4} \\
\\
\tau_1 \otimes \tau_2 &=& \tau_1 * \tau_2 \\
\tau \otimes (\nodet{\cubt_1}{\cubt_2}) &=& 
  \nodet{\tau \otimes \cubt_1}{\tau \otimes \cubt_2} \\
(\nodet{\cubt_1}{\cubt_2}) \otimes \cubt &=& 
  \nodet{\cubt_1 \otimes \cubt}{\cubt_2 \otimes \cubt}
\end{array}\]

\begin{figure*}
\[\begin{array}{c}
\nodet{\tau_1}{\tau_2}
\quad\otimes\quad
\nodet{(\nodet{\tau_3}{\tau_4})}{(\nodet{\tau_5}{\tau_6})} = 
\\
\nodet{(\nodet{{(\nodet{\tau_1*\tau_3}{\tau_1*\tau_4})}}
              {{(\nodet{\tau_1*\tau_5}{\tau_1*\tau_6})}})}
      {(\nodet{{(\nodet{\tau_2*\tau_3}{\tau_2*\tau_4})}}
              {{(\nodet{\tau_2*\tau_5}{\tau_2*\tau_6})}})}
\end{array}\]
\caption{\label{mult}Example of multiplication of two cubical types.}
\end{figure*}

A few examples might provide some needed intuition. First we note that the
general definitions of sum and product reduce to the conventional sums and
products for 1-dimensional types. Also the sum of 2-dimensional types reduces
to the sum used in the \textbf{Int} construction. The interesting new
development is the product definition. The example in Fig.~\ref{mult} shows
that in general the product of an $n$-dimensional type with an
$m$-dimensional type gives an $n+m$-dimensional type. For low dimensions such
as the ones used in the example above, it is possible to visualize what is
happening in coordinate space.

\begin{figure*}
\begin{tikzpicture}
\node[left] at (0,0) {$\tau_1^+$};
\node[right] at (0.6,0) {$\tau_2^-$};
\draw[-] (0,0) -- (0.6,0);
\node at (1.6,0) {$\otimes$}; 
%%
\node[below left] at (2.5,-0.5) {$\tau_3^+$};
\node[below right] at (3.5,-0.5) {$\tau_4^-$};
\draw[-] (2.5,-0.5) -- (3.5,-0.5);
\draw[-] (2.5,-0.5) -- (2.5,0.5);
\node[above left] at (2.5,0.5) {$\tau_6^-$};
\node[above right] at (3.5,0.5) {$\tau_5^+$};
\draw[-] (2.5,0.5) -- (3.5,0.5);
\draw[-] (3.5,-0.5) -- (3.5,0.5);
%% 
\node at (4.5,0) {$=$};
%% 
\node[left] at (7.5,0.75) {$(\tau_2*\tau_3)^-$};
\node[right] at (9.5,0.75) {$(\tau_2*\tau_4)^+$};
\node[above right] at (10.2,1.2) {$(\tau_2*\tau_5)^-$};
\node[above left] at (8.2,1.2) {$(\tau_2*\tau_6)^+$};
%%
\node[left] at (7.5,-0.75) {$(\tau_1*\tau_3)^+$};
\node[right] at (9.5,-0.75) {$(\tau_1*\tau_4)^-$};
\node[above right] at (10.2,-0.3) {$(\tau_1*\tau_5)^+$};
\node[above] at (8.2,-0.3) {$(\tau_1*\tau_6)^-$};
%%
\draw[-] (7.5,0.75) -- (9.5,0.75);
\draw[-] (9.5,0.75) -- (10.2,1.2);
\draw[-] (10.2,1.2) -- (8.2,1.2);
\draw[-] (8.2,1.2) -- (7.5,0.75);
%%
\draw[-] (7.5,-0.75) -- (9.5,-0.75);
\draw[-] (9.5,-0.75) -- (10.2,-0.3);
\draw[-,dashed] (10.2,-0.3) -- (8.2,-0.3);
\draw[-,dashed] (8.2,-0.3) -- (7.5,-0.75);
%%
\draw[-] (7.5,0.75) -- (7.5,-0.75);
\draw[-] (9.5,0.75) -- (9.5,-0.75);
\draw[-] (10.2,1.2) -- (10.2,-0.3);
\draw[-,dashed] (8.2,1.2) -- (8.2,-0.3);
\end{tikzpicture}
\caption{\label{multv}Visualization of multiplication of two cubical types.}
\end{figure*}

\begin{verbatim}
There are zero objects of all depths:
  0, (0-0), (0-0)-0, ...
There are 1 objects of all depths:
  1, (1-0), (1-0)-0, ...

To complete the story we need to 
define morphisms. (More on this below.)
Once we have a notion of morphism we 
can check whether X + 0 is the same
as X etc. i.e., we can check all the 
ring equivalences.

Ok so what are the morphisms between 
these cubical objects? We know what
they are for 1-dimensional cubes: 
they are the pi combinabors. We also
know what they are for the 2-dimensional 
cubes: a maps (A-B) ==> (C-D) 
is a Pi map between A+D <=> C+B. 
How to generalize this?

Why is there no trace in the ring 
completion paper??? What are 
the morphisms in that paper?
\end{verbatim}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{A Reversible Language with Cubical Types}

Pi with cubical types

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Related Work and Context}

A ton of stuff here.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusion}
